---
title: "programming"
author: "Hernán Estrin"
date: "8/10/2019"
output: html_document
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this part of the book, you’ll improve your programming skills. Programming is a cross-cutting skill needed for all data science work: you must use a computer to do data science; you cannot do it in your head, or with pencil and paper.

Programming produces code, and code is a tool of communication. Obviously code tells the computer what you want it to do. But it also communicates meaning to other humans. Thinking about code as a vehicle for communication is important because every project you do is fundamentally collaborative. Even if you’re not working with other people, you’ll definitely be working with future-you! Writing clear code is important so that others (like future-you) can understand why you tackled an analysis in the way you did. That means getting better at programming also involves getting better at communicating. Over time, you want your code to become not just easier to write, but easier for others to read.

Writing code is similar in many ways to writing prose. One parallel which I find particularly useful is that in both cases rewriting is the key to clarity. The first expression of your ideas is unlikely to be particularly clear, and you may need to rewrite multiple times. After solving a data analysis challenge, it’s often worth looking at your code and thinking about whether or not it’s obvious what you’ve done. If you spend a little time rewriting your code while the ideas are fresh, you can save a lot of time later trying to recreate what your code did. But this doesn’t mean you should rewrite every function: you need to balance what you need to achieve now with saving time in the long run. 

In the following four chapters, you’ll learn skills that will allow you to both tackle new programs and to solve existing problems with greater clarity and ease:

* In *pipes*, you will dive deep into the pipe, %>%, and learn more about how it works, what the alternatives are, and when not to use it.

* Copy-and-paste is a powerful tool, but you should avoid doing it more than twice. Repeating yourself in code is dangerous because it can easily lead to errors and inconsistencies. Instead, in *functions*, you’ll learn how to write functions which let you extract out repeated code so that it can be easily reused.

* As you start to write more powerful functions, you’ll need a solid grounding in *R’s data structures*, provided by vectors. You must master the four common atomic vectors, the three important S3 classes built on top of them, and understand the mysteries of the list and data frame.

* Functions extract out repeated code, but you often need to repeat the same actions on different inputs. You need tools for iteration that let you do similar things again and again. These tools include *for loops and functional programming*, which you’ll learn about in iteration.

# Pipes

Pipes are a powerful tool for clearly expressing a sequence of multiple operations. So far, you’ve been using them without knowing how they work, or what the alternatives are. Now, in this chapter, it’s time to explore the pipe in more detail. You’ll learn the alternatives to the pipe, when you shouldn’t use the pipe, and some useful related tools.

The pipe, %>%, comes from the magrittr package by Stefan Milton Bache. Packages in the tidyverse load %>% for you automatically, so you don’t usually load magrittr explicitly. Here, however, we’re focussing on piping, and we aren’t loading any other packages, so we will load it explicitly.

```{r loading magrittr}
library(magrittr)
```

Pipe is my favourite form, because it focusses on verbs, not nouns. You can read this series of function compositions like it’s a set of imperative actions. Foo Foo hops, then scoops, then bops. The downside, of course, is that you need to be familiar with the pipe. If you’ve never seen %>% before, you’ll have no idea what this code does. Fortunately, most people pick up the idea very quickly, so when you share your code with others who aren’t familiar with the pipe, you can easily teach them.

The pipe works by performing a “lexical transformation”: behind the scenes, magrittr reassembles the code in the pipe to a form that works by overwriting an intermediate object.

Pipe won’t work for two classes of functions:

* Functions that use the current environment. For example, assign() will create a new variable with the given name in the current environment. The use of assign with the pipe does not work because it assigns it to a temporary environment used by %>%. If you do want to use assign with the pipe, you must be explicit about the environment:

```{r pipe assignation}
env <- environment()
"x" %>% assign(100, envir = env)
x
```

Other functions with this problem include get() and load().

* Functions that use lazy evaluation. In R, function arguments are only computed when the function uses them, not prior to calling the function. The pipe computes each element in turn, so you can’t rely on this behaviour.

The pipe is a powerful tool, but it’s not the only tool at your disposal, and it doesn’t solve every problem! Pipes are most useful for rewriting a fairly short linear sequence of operations. I think you should reach for another tool when:

* Your pipes are longer than (say) ten steps. In that case, create intermediate objects with meaningful names. That will make debugging easier, because you can more easily check the intermediate results, and it makes it easier to understand your code, because the variable names can help communicate intent.

* You have multiple inputs or outputs. If there isn’t one primary object being transformed, but two or more objects being combined together, don’t use the pipe.

* You are starting to think about a directed graph with a complex dependency structure. Pipes are fundamentally linear and expressing complex relationships with them will typically yield confusing code.

## Other tools from magrittr

All packages in the tidyverse automatically make %>% available for you, so you don’t normally load magrittr explicitly. However, there are some other useful tools inside magrittr that you might want to try out:

* When working with more complex pipes, it’s sometimes useful to call a function for its side-effects. Maybe you want to print out the current object, or plot it, or save it to disk. Many times, such functions don’t return anything, effectively terminating the pipe.

To work around this problem, you can use the “tee” pipe. %T>% works like %>% except that it returns the left-hand side instead of the right-hand side. It’s called “tee” because it’s like a literal T-shaped pipe.

```{r}
rnorm(100) %>%
  matrix(ncol = 2) %T>%
  plot() %>%
  str()
```

* If you’re working with functions that don’t have a data frame based API (i.e. you pass them individual vectors, not a data frame and expressions to be evaluated in the context of that data frame), you might find %$% useful. It “explodes” out the variables in a data frame so that you can refer to them explicitly. This is useful when working with many functions in base R:

```{r}
mtcars %$%
  cor(disp, mpg)
```


* For assignment magrittr provides the %<>% operator wh

```{r}
mtcars %<>% transform(cyl = cyl * 2)
```

I’m not a fan of this operator because I think assignment is such a special operation that it should always be clear when it’s occurring. In my opinion, a little bit of duplication (i.e. repeating the name of the object twice) is fine in return for making assignment more explicit.

# Functions

## Generalities

One of the best ways to improve your reach as a data scientist is to write functions. Functions allow you to automate common tasks in a more powerful and general way than copy-and-pasting. Writing a function has three big advantages over using copy-and-paste:

* You can give a function an evocative name that makes your code easier to understand.

* As requirements change, you only need to update code in one place, instead of many.

* You eliminate the chance of making incidental mistakes when you copy and paste (i.e. updating a variable name in one place, but not in another).

Writing good functions is a lifetime journey. Even after using R for many years I still learn new techniques and better ways of approaching old problems. The goal of this chapter is not to teach you every esoteric detail of functions but to get you started with some pragmatic advice that you can apply immediately.

As well as practical advice for writing functions, this chapter also gives you some suggestions for how to style your code. Good code style is like correct punctuation. Youcanmanagewithoutit, but it sure makes things easier to read! As with styles of punctuation, there are many possible variations. Here we present the style we use in our code, but the most important thing is to be consistent.

You should consider writing a function whenever you’ve copied and pasted a block of code more than twice (i.e. you now have three copies of the same code).

To write a function you need to first analyse the code. How many inputs does it have? To make the inputs more clear, it’s a good idea to rewrite the code using temporary variables with general names.

There are three key steps to creating a new function:

* You need to pick a name for the function. Here I’ve used rescale01 because this function rescales a vector to lie between 0 and 1.

* You list the inputs, or arguments, to the function inside function. Here we have just one argument. If we had more the call would look like function(x, y, z).

* You place the code you have developed in body of the function, a { block that immediately follows function(...).

```{r rescale function}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
```


Note the overall process: I only made the function after I’d figured out how to make it work with a simple input. It’s easier to start with working code and turn it into a function; it’s harder to create a function and then try to make it work.

At this point it’s a good idea to check your function with a few different inputs. As you write more and more functions you’ll eventually want to convert these informal, interactive tests into formal, automated tests. That process is called unit testing. 

Compared to the original, this code is easier to understand and we’ve eliminated one class of copy-and-paste errors.

Another advantage of functions is that if our requirements change, we only need to make the change in one place.  This is an important part of the “do not repeat yourself” (or DRY) principle. The more repetition you have in your code, the more places you need to remember to update when things change (and they always do!), and the more likely you are to create bugs over time.

## Good practices 

It’s important to remember that functions are not just for the computer, but are also for humans. R doesn’t care what your function is called, or what comments it contains, but these are important for human readers. This section discusses some things that you should bear in mind when writing functions that humans can understand.

The name of a function is important. Ideally, the name of your function will be short, but clearly evoke what the function does. That’s hard! But it’s better to be clear than short, as RStudio’s autocomplete makes it easy to type long names.

Generally, function names should be verbs, and arguments should be nouns. There are some exceptions: nouns are ok if the function computes a very well known noun. A good sign that a noun might be a better choice is if you’re using a very broad verb like “get”, “compute”, “calculate”, or “determine”. Use your best judgement and don’t be afraid to rename a function if you figure out a better name later.


```{r}
# Too short
f()

# Not a verb, or descriptive
my_awesome_function()

# Long, but clear
impute_missing()
collapse_years()
```


If your function name is composed of multiple words, I recommend using “snake_case”, where each lowercase word is separated by an underscore. camelCase is a popular alternative. It doesn’t really matter which one you pick, the important thing is to be consistent: pick one or the other and stick with it. R itself is not very consistent, but there’s nothing you can do about that. Make sure you don’t fall into the same trap by making your code as consistent as possible.

```{r}
# Never do this!
col_mins <- function(x, y) {}
rowMaxes <- function(y, x) {}
```


If you have a family of functions that do similar things, make sure they have consistent names and arguments. Use a common prefix to indicate that they are connected. That’s better than a common suffix because autocomplete allows you to type the prefix and see all the members of the family.

```{r}
# Good
input_select()
input_checkbox()
input_text()

# Not so good
select_input()
checkbox_input()
text_input()
```


Where possible, avoid overriding existing functions and variables. It’s impossible to do in general because so many good names are already taken by other packages, but avoiding the most common names from base R will avoid confusion.

```{r}
# Don't do this!
T <- FALSE
c <- 10
mean <- function(x) sum(x)
```


Use comments, lines starting with #, to explain the “why” of your code. You generally should avoid comments that explain the “what” or the “how”. If you can’t understand what the code does from reading it, you should think about how to rewrite it to be more clear. Do you need to add some intermediate variables with useful names? Do you need to break out a subcomponent of a large function so you can name it? -sions: why did you choose this approach instead of an alternative? What else did you try that didn’t work? It’s a great idea to capture that sort of thinking in a comment.

Another important use of comments is to break up your file into easily readable chunks. Use long lines of - and = to make it easy to spot the breaks. RStudio provides a keyboard shortcut to create these headers (Cmd/Ctrl + Shift + R), and will display them in the code navigation drop-down at the bottom-left of the editor.


```{r}
# Load data --------------------------------------

# Plot data --------------------------------------
```

## Conditional execution


An `r if` statement allows you to conditionally execute code. It looks like this:

```{r if statement}
if (condition) {
  # code executed when condition is TRUE
} else {
  # code executed when condition is FALSE
}
```

Here’s a simple function that uses an if statement. The goal of this function is to return a logical vector describing whether or not each element of a vector is named.

```{r if example}
has_name <- function(x) {
  nms <- names(x)
  if (is.null(nms)) {
    rep(FALSE, length(x))
  } else {
    !is.na(nms) & nms != ""
  }
}
```

This function takes advantage of the standard return rule: a function returns the last value that it computed. Here that is either one of the two branches of the if statement.

### Conditions

The condition must evaluate to either TRUE or FALSE. If it’s a vector, you’ll get a warning message; if it’s an NA, you’ll get an error. Watch out for these messages in your own code:

```{r}
if (c(TRUE, FALSE)) {}

if (NA) {}

```

You can use || (or) and && (and) to combine multiple logical expressions. These operators are “short-circuiting”: as soon as || sees the first TRUE it returns TRUE without computing anything else. As soon as && sees the first FALSE it returns FALSE. You should never use | or & in an if statement: these are vectorised operations that apply to multiple values (that’s why you use them in filter()). If you do have a logical vector, you can use any() or all() to collapse it to a single value.

Be careful when testing for equality. == is vectorised, which means that it’s easy to get more than one output. Either check the length is already 1, collapse with all() or any(), or use the non-vectorised identical(). identical() is very strict: it always returns either a single TRUE or a single FALSE, and doesn’t coerce types. This means that you need to be careful when comparing integers and doubles:

```{r identical}
identical(0L, 0)
```

You also need to be wary of floating point numbers:

```{r floating points error}
x <- sqrt(2) ^ 2
x
x == 2
x - 2
dplyr::near(x, 2)
```

Instead use dplyr::near() for comparisons, as described in comparisons.

And remember, x == NA doesn’t do anything useful!

## Multiple conditions

You can chain multiple if statements together:

```{r multiple conditions}
if (this) {
  # do that
} else if (that) {
  # do something else
} else {
  # 
}
```

But if you end up with a very long series of chained if statements, you should consider rewriting. One useful technique is the switch() function. It allows you to evaluate selected code based on position or name.

```{r}
fx <- 
  function(x, y, op) {
  switch(op,
         plus = x + y,
         minus = x - y,
         times = x * y,
         divide = x / y,
         stop("Unknown op!")
  )
  }

fx(4,2, "times")
```

## Code style

Both if and function should (almost) always be followed by squiggly brackets ({}), and the contents should be indented by two spaces. This makes it easier to see the hierarchy in your code by skimming the left-hand margin.

An opening curly brace should never go on its own line and should always be followed by a new line. A closing curly brace should always go on its own line, unless it’s followed by else. Always indent the code inside curly braces.

```{r code style}
# Good
if (y < 0 && debug) {
  message("Y is negative")
}

if (y == 0) {
  log(x)
} else {
  y ^ x
}

# Bad
if (y < 0 && debug)
message("Y is negative")

if (y == 0) {
  log(x)
} 
else {
  y ^ x
}
```

It’s ok to drop the curly braces if you have a very short if statement that can fit on one line:

```{r}
y <- 10
x <- if (y < 20) "Too low" else "Too high"
```

I recommend this only for very brief if statements. Otherwise, the full form is easier to read:

```{r}
if (y < 20) {
  x <- "Too low" 
} else {
  x <- "Too high"
}
```

## Function arguments

The arguments to a function typically fall into two broad sets: one set supplies the data to compute on, and the other supplies arguments that control the details of the computation. 

Generally, data arguments should come first. Detail arguments should go on the end, and usually should have default values. You specify a default value in the same way you call a function with a named argument:

```{r}
mean_ci <- function(x, conf = 0.95) {
  se <- sd(x) / sqrt(length(x))
  alpha <- 1 - conf
  mean(x) + se * qnorm(c(alpha / 2, 1 - alpha / 2))
}

x <- runif(100)
mean_ci(x)
mean_ci(x, conf = 0.99)

```

The default value should almost always be the most common value. The few exceptions to this rule are to do with safety. For example, it makes sense for na.rm to default to FALSE because missing values are important. Even though na.rm = TRUE is what you usually put in your code, it’s a bad idea to silently ignore missing values by default.

When you call a function, you typically omit the names of the data arguments, because they are used so commonly. If you override the default value of a detail argument, you should use the full name.

```{r}
# Good
mean(1:10, na.rm = TRUE)

# Bad
mean(x = 1:10, , FALSE)
mean(, TRUE, x = c(1:10, NA))
```

You can refer to an argument by its unique prefix (e.g. mean(x, n = TRUE)), but this is generally best avoided given the possibilities for confusion.

Notice that when you call a function, you should place a space around = in function calls, and always put a space after a comma, not before (just like in regular English). Using whitespace makes it easier to skim the function for the important components.

```{r}
# Good
average <- mean(feet / 12 + inches, na.rm = TRUE)

# Bad
average<-mean(feet/12+inches,na.rm=TRUE)
```

### Choosing names

The names of the arguments are also important. R doesn’t care, but the readers of your code (including future-you!) will. Generally you should prefer longer, more descriptive names, but there are a handful of very common, very short names. It’s worth memorising these:

* x, y, z: vectors.
* w: a vector of weights.
* df: a data frame.
* i, j: numeric indices (typically rows and columns).
* n: length, or number of rows.
* p: number of columns.

Otherwise, consider matching names of arguments in existing R functions. For example, use na.rm to determine if missing values should be removed.


### Checking values

As you start to write more functions, you’ll eventually get to the point where you don’t remember exactly how your function works. At this point it’s easy to call your function with invalid inputs. To avoid this problem, it’s often useful to make constraints explicit. 

It’s good practice to check important preconditions, and throw an error (with stop()), if they are not true:

```{r}
wt_mean <- function(x, w) {
  if (length(x) != length(w)) {
    stop("hey,`x` and `w` must be the same length", call. = FALSE)
  }
  sum(w * x) / sum(w)
}

wt_mean(c(2,3), 1)
```

Be careful not to take this too far. There’s a tradeoff between how much time you spend making your function robust, versus how long you spend writing it. For example, if you also added a na.rm argument, I probably wouldn’t check it carefully. 

This is a lot of extra work for little additional gain. A useful compromise is the built-in stopifnot(): it checks that each argument is TRUE, and produces a generic error message if not.

```{r}
wt_mean <- function(x, w, na.rm = FALSE) {
  stopifnot(is.logical(na.rm), length(na.rm) == 1)
  stopifnot(length(x) == length(w))
  
  if (na.rm) {
    miss <- is.na(x) | is.na(w)
    x <- x[!miss]
    w <- w[!miss]
  }
  sum(w * x) / sum(w)
}

wt_mean(1:6, 6:1, na.rm = "ss") 
wt_mean(1:4, 6:1, na.rm = F) 
```

Note that when using stopifnot() you assert what should be true rather than checking for what might be wrong.

###  Dot-dot-dot (…)

Many functions in R take an arbitrary number of inputs. 

How do these functions work? They rely on a special argument: ... (pronounced dot-dot-dot). This special argument captures any number of arguments that aren’t otherwise matched.

It’s useful because you can then send those ... on to another function. This is a useful catch-all if your function primarily wraps another function. For example, I commonly create these helper functions that 

```{r dot dot dot}
commas <- function(...) stringr::str_c(..., collapse = ", ")
commas(letters[1:10])
#> [1] "a, b, c, d, e, f, g, h, i, j"

rule <- function(..., pad = "-") {
  title <- paste0(...)
  width <- getOption("width") - nchar(title) - 5
  cat(title, " ", stringr::str_dup(pad, width), "\n", sep = "")
}

rule("Important output", " here")
```

Here ... lets me forward on any arguments that I don’t want to deal with to str_c(). It’s a very convenient technique. But it does come at a price: any misspelled arguments will not raise an error. This makes it easy for typos to go unnoticed:

```{r}
x <- c(1, 2)
sum(x, na.mr = TRUE) # Adds 1 for True
```

If you just want to capture the values of the ..., use list(...).

```{r}
capture  <- function(...) list(...)
capture(1, 2, 3)
```

### Lazy evaluation

Arguments in R are lazily evaluated: they’re not computed until they’re needed. That means if they’re never used, they’re never called. This is an important property of R as a programming language, but is generally not important when you’re writing your own functions for data analysis. 


## Return values

Figuring out what your function should return is usually straightforward: it’s why you created the function in the first place! There are two things you should consider when returning a value:

1. Does returning early make your function easier to read?
2. Can you make your function pipeable?

### Explicit return statements

The value returned by the function is usually the last statement it evaluates, but you can choose to return early by using return(). I think it’s best to save the use of return() to signal that you can return early with a simpler solution. A common reason to do this is because the inputs are empty:

```{r early return}
complicated_function <- function(x, y, z) {
  if (length(x) == 0 || length(y) == 0) {
    return(0)
  }
 # Complicated code here
}
```


Another reason is because you have a if statement with one complex block and one simple block.  If the first block is very long, by the time you get to the else, you’ve forgotten the condition. One way to rewrite it is to use an early return for the simple case:

```{r early return 2}
f <- function() {
  if (!x) {
    return(something_short)
  }

  # Do 
  # something
  # that
  # takes
  # many
  # lines
  # to
  # express
}
```

This tends to make the code easier to understand, because you don’t need quite so much context to understand it.

### Writing pipeable functions

If you want to write your own pipeable functions, it’s important to think about the return value. Knowing the return value’s object type will mean that your pipeline will “just work”. For example, with dplyr and tidyr the object type is the data frame.

There are two basic types of pipeable functions: *transformations* and *side-effects.* With transformations, an object is passed to the function’s first argument and a modified object is returned. With side-effects, the passed object is not transformed. Instead, the function performs an action on the object, like drawing a plot or saving a file. Side-effects functions should “invisibly” return the first argument, so that while they’re not printed they can still be used in a pipeline. 

```{r invisibly return object}
show_missings <- function(df) {
  n <- sum(is.na(df))
  cat("Missing values: ", n, "\n", sep = "")
  invisible(df)
}

show_missings(mtcars)
```

If we call it interactively, the invisible() means that the input df doesn’t get printed out. But it’s still there, it’s just not printed by default.

```{r}
x <- show_missings(mtcars) 
class(x)
dim(x)
```

And we can still use it in a pipe:

```{r}
mtcars %>% 
  show_missings() %>% 
  mutate(mpg = ifelse(mpg < 20, NA, mpg)) %>% 
  show_missings() 
```

## Environment

The last component of a function is its environment. This is not something you need to understand deeply when you first start writing functions. However, it’s important to know a little bit about environments because they are crucial to how functions work. The environment of a function controls how R finds the value associated with a name. For example, take this function:

```{r}
f <- function(x) {
  x + y
}
```

In many programming languages, this would be an error, because y is not defined inside the function. In R, this is valid code because R uses rules called lexical scoping to find the value associated with a name. Since y is not defined inside the function, R will look in the environment where the function was defined:

```{r}
y <- 100
f(10)
y <- 1000
f(10)
```

This behaviour seems like a recipe for bugs, and indeed you should avoid creating functions like this deliberately, but by and large it doesn’t cause too many problems (especially if you regularly restart R to get to a clean slate).

The advantage of this behaviour is that from a language standpoint it allows R to be very consistent. Every name is looked up using the same set of rules.