---
title: "strings"
author: "Hernán Estrin"
date: "August 8, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
```

This chapter introduces you to string manipulation in R. You’ll learn the basics of how strings work and how to create them by hand, but the focus of this chapter will be on regular expressions, or regexps for short. Regular expressions are useful because strings usually contain unstructured or semi-structured data, and regexps are a concise language for describing patterns in strings. When you first look at a regexp, you’ll think a cat walked across your keyboard, but as your understanding improves they will soon start to make sense.

# String basics

You can create strings with either single quotes or double quotes. Unlike other languages, there is no difference in behaviour. I recommend always using ", unless you want to create a string that contains multiple ".

```{r create strings}
string1 <- "This is a string"
string2 <- 'If I want to include a "quote" inside a string, I use single quotes'
```

To include a literal single or double quote in a string you can use \ to “escape” it:

```{r escaping quotes}
double_quote <- "\"" # or '"'
single_quote <- '\'' # or "'"
backslash <- "\\"
```

Beware that the printed representation of a string is not the same as string itself, because the printed representation shows the escapes. To see the raw contents of the string, use writeLines():

```{r see raw content}
writeLines(backslash)
writeLines(double_quote)
writeLines(single_quote)
```

There are a handful of other special characters. The most common are "\n", newline, and "\t", tab, but you can see the complete list by requesting help on ": ?'"', or ?"'". You’ll also sometimes see strings like "\u00b5", this is a way of writing non-English characters that works on all platforms:

```{r unicode}
(x <- "\u00b5")
```

Multiple strings are often stored in a character vector, which you can create with c():

```{r character vector}
c("one", "two", "three")
```

## String length

Base R contains many functions to work with strings but we’ll avoid them because they can be inconsistent, which makes them hard to remember. Instead we’ll use functions from stringr. These have more intuitive names, and all start with str_. For example, str_length() tells you the number of characters in a string:

```{r str length}
str_length(c("a", "R for data science", NA))
```

The common str_ prefix is particularly useful if you use RStudio, because typing str_ will trigger autocomplete, allowing you to see all stringr functions.

## Combining strings

To combine two or more strings, use str_c():

```{r combine strings}
str_c("x", "y")

str_c("x", "y", "z")
```

Use the sep argument to control how they’re separated:

```{r combine strings 2}
str_c("x", "y", sep = "/ ")
```

Like most other functions in R, missing values are contagious. If you want them to print as "NA", use str_replace_na():

```{r combine NA strings}
x <- c("abc", NA)
str_c("|-", x, "-|")
str_c("|-", str_replace_na(x), "-|") ## Turn NA into "NA"
```

As shown above, str_c() is vectorised, and it automatically recycles shorter vectors to the same length as the longest:

```{r recycling vectors when combining}
str_c("prefix-", c("a", "b", "c"), "-suffix")
```

Objects of length 0 are silently dropped. This is particularly useful in conjunction with if:

```{r dropping 0s}
name <- "Hadley"
time_of_day <- "morning"
birthday <- FALSE

str_c(
  "Good ", time_of_day, " ", name,
  if (birthday) " and HAPPY BIRTHDAY", ## Not run cause it's FALSE
  "."
)
```

To collapse a vector of strings into a single string, use collapse:

```{r collapse vector of strings}
str_c(c("x", "y", "z"), collapse = ", ")
```

## Subsetting strings

You can extract parts of a string using str_sub(). As well as the string, str_sub() takes start and end arguments which give the (inclusive) position of the substring:

```{r substrings}
x <- c("Apple", "Banana", "Pear")
str_sub(x, 1, str_length(x)-2)
str_sub(x, str_length(x)-1, str_length(x))
```

Note that str_sub() won’t fail if the string is too short: it will just return as much as possible:

```{r str_sub return as much as possible}
str_sub("a", 1, 5)
```

You can also use the assignment form of str_sub() to modify strings:

```{r substr to modify strings}
(str_sub(x, 1, 1) <- str_to_lower(str_sub(x, 1, 1))) # Making first letter lower-case
```

## Locales

Above I used str_to_lower() to change the text to lower case. You can also use str_to_upper() or str_to_title(). However, changing case is more complicated than it might at first appear because different languages have different rules for changing case. You can pick which set of rules to use by specifying a locale:

```{r to_upper}
str_to_upper(c("i", "ı"))
str_to_upper(c("i", "ı"), locale = "tr")
```


The locale is specified as a ISO 639 language code, which is a two or three letter abbreviation. If you don’t already know the code for your language, Wikipedia has a good list. If you leave the locale blank, it will use the current locale, as provided by your operating system.

Another important operation that’s affected by the locale is sorting. The base R order() and sort() functions sort strings using the current locale. If you want robust behaviour across different computers, you may want to use str_sort() and str_order() which take an additional locale argument:

```{r sorting using locale}
x <- c("apple", "eggplant", "banana")
str_sort(x, locale = "en") 
str_sort(x, locale = "haw")
```

# Matching patterns with regular expressions

Regexps are a very terse language that allow you to describe patterns in strings. They take a little while to get your head around, but once you understand them, you’ll find them extremely useful.

To learn regular expressions, we’ll use str_view() and str_view_all(). These functions take a character vector and a regular expression, and show you how they match. We’ll start with very simple regular expressions and then gradually get more and more complicated. Once you’ve mastered pattern matching, you’ll learn how to apply those ideas with various stringr functions.

## Basics

The simplest patterns match exact strings:

```{r basic match}
x <- c("apple", "banana", "pear")
str_view(x, "an")
```

The next step up in complexity is ., which matches any character (except a newline):

```{r dot in regex}
str_view(x, ".a.")
```

But if “.” matches any character, how do you match the character “.”? You need to use an “escape” to tell the regular expression you want to match it exactly, not use its special behaviour. Like strings, regexps use the backslash, \, to escape special behaviour. So to match an ., you need the regexp \.. Unfortunately this creates a problem. We use strings to represent regular expressions, and \ is also used as an escape symbol in strings. So to create the regular expression \. we need the string "\\.".

```{r escaping characters}
dot <- "\\."

writeLines(dot)

str_view(c("abc", "a.c", "bef"), "a\\.c")
```

If \ is used as an escape character in regular expressions, how do you match a literal \? Well you need to escape it, creating the regular expression \\. To create that regular expression, you need to use a string, which also needs to escape \. That means to match a literal \ you need to write "\\\\" — you need four backslashes to match one!

```{r escaping backslash}
x <- "a\\b"
writeLines(x)
str_view(x, "\\\\")
```

In this book, I’ll write regular expression as \. and strings that represent the regular expression as "\\.".

## Anchors

By default, regular expressions will match any part of a string. It’s often useful to anchor the regular expression so that it matches from the start or end of the string. You can use:

* ^ to match the start of the string.
* $ to match the end of the string.

```{r anchors}
x <- c("apple", "banana", "pear")
str_view(x, "^a")
str_view(x, "a$")
```

To force a regular expression to only match a complete string, anchor it with both ^ and $:

```{r complete string}
x <- c("apple pie", "apple", "apple cake")
str_view(x, "apple")
str_view(x, "^apple$")
```

You can also match the boundary between words with \b. I don’t often use this in R, but I will sometimes use it when I’m doing a search in RStudio when I want to find the name of a function that’s a component of other functions.

## Character classes and alternatives

There are a number of special patterns that match more than one character. You’ve already seen ., which matches any character apart from a newline. There are four other useful tools:

* \d: matches any digit.
* \s: matches any whitespace (e.g. space, tab, newline).
* [abc]: matches a, b, or c.
* [^abc]: matches anything except a, b, or c.

Remember, to create a regular expression containing \d or \s, you’ll need to escape the \ for the string, so you’ll type "\\d" or "\\s".

A character class containing a single character is a nice alternative to backslash escapes when you want to include a single metacharacter in a regex. Many people find this more readable.

```{r character classes}
# Look for a literal character that normally has special meaning in a regex
str_view(c("abc", "a.c", "a*c", "a c"), "a[.]c")
str_view(c("abc", "a.c", "a*c", "a c"), ".[*]c")
str_view(c("abc", "a.c", "a*c", "a c"), "a[ ]")
str_view(c("abc", "a8c", "a*c", "a c"), "a\\s")
str_view(c("abc", "a8c", "a*c", "a c"), "\\d")
```

This works for most (but not all) regex metacharacters: $ . | ? * + ( ) [ {. Unfortunately, a few characters have special meaning even inside a character class and must be handled with backslash escapes: ] \ ^ and -.

You can use alternation to pick between one or more alternative patterns. For example, abc|d..f will match either ‘“abc”’, or "deaf". Note that the precedence for | is low, so that abc|xyz matches abc or xyz not abcyz or abxyz. Like with mathematical expressions, if precedence ever gets confusing, use parentheses to make it clear what you want:

```{r alternation}
str_view(c("abc","xyz","abcyz","abxyz"), "abc|xyz")
str_view(c("abc","xyz","abcyz","abxyz"), "ab(c|x)yz")
```

## Repetition

The next step up in power involves controlling how many times a pattern matches:

* ?: 0 or 1
* +: 1 or more
* *: 0 or more

```{r repetition patterns}
x <- c("ab","abc","ac","abbc","abd")
str_view(x, "ab?c")
str_view(x, "ab+c")
str_view(x, "ab*c")
```

Note that the precedence of these operators is high, so you can write: colou?r to match either American or British spellings. That means most uses will need parentheses, like bana(na)+.

You can also specify the number of matches precisely:

* {n}: exactly n
* {n,}: n or more
* {,m}: at most m
* {n,m}: between n and m

```{r repetition patterns 2}
x <- c("abc","abbc","abbbc","abbbbc","abbbbbc")
str_view(x, "ab{2}c")
str_view(x, "ab{2,}c")
str_view(x, "ab{1,3}c")
str_view(x, "ab{2,4}c")
```

## Grouping and backreferences

Earlier, you learned about parentheses as a way to disambiguate complex expressions. Parentheses also create a numbered capturing group (number 1, 2 etc.). A capturing group stores the part of the string matched by the part of the regular expression inside the parentheses. You can refer to the same text as previously matched by a capturing group with backreferences, like \1, \2 etc.

```{r backreference}
str_view(fruit, "(..)\\1", match = TRUE)
```


# Tools

Now that you’ve learned the basics of regular expressions, it’s time to learn how to apply them to real problems. In this section you’ll learn a wide array of stringr functions that let you:

* Determine which strings match a pattern.
* Find the positions of matches.
* Extract the content of matches.
* Replace matches with new values.
* Split a string based on a match.

A word of caution before we continue: because regular expressions are so powerful, it’s easy to try and solve every problem with a single regular expression.

Don’t forget that you’re in a programming language and you have other tools at your disposal. Instead of creating one complex regular expression, it’s often easier to write a series of simpler regexps. If you get stuck trying to create a single regexp that solves your problem, take a step back and think if you could break the problem down into smaller pieces, solving each challenge before moving onto the next one.

## Detect matches

To determine if a character vector matches a pattern, use str_detect(). It returns a logical vector the same length as the input:

```{r str detect}
x <- c("apple", "banana", "pear")
str_detect(x, "e")
```

Remember that when you use a logical vector in a numeric context, FALSE becomes 0 and TRUE becomes 1. That makes sum() and mean() useful if you want to answer questions about matches across a larger vector:

```{r counting and perc total matching}
sum(str_detect(words, "^t"))
mean(str_detect(words, "[aeiou]$"))
```

When you have complex logical conditions (e.g. match a or b but not c unless d) it’s often easier to combine multiple str_detect() calls with logical operators, rather than trying to create a single regular expression. For example, here are two ways to find all words that don’t contain any vowels:

```{r simplifying regex}
no_vowels_1 <- !str_detect(words, "[aeiou]")
no_vowels_2 <- str_detect(words, "^[^aeiou]+$")
identical(no_vowels_1, no_vowels_2)
```

The results are identical, but I think the first approach is significantly easier to understand. If your regular expression gets overly complicated, try breaking it up into smaller pieces, giving each piece a name, and then combining the pieces with logical operations.

A common use of str_detect() is to select the elements that match a pattern. You can do this with logical subsetting, or the convenient str_subset() wrapper:

```{r subsetting}
words[str_detect(words, "x$")]
str_subset(words, "x$")
```

Typically, however, your strings will be one column of a data frame, and you’ll want to use filter instead:

```{r filter dataframe with regex}
df <- tibble(
  word = words, 
  i = seq_along(word)
)
df %>% 
  filter(str_detect(word, "x$"))
```

A variation on str_detect() is str_count(): rather than a simple yes or no, it tells you how many matches there are in a string:

```{r counting number of matchings}
x <- c("apple", "banana", "pear")
str_count(x, "a")
mean(str_count(words, "[aeiou]"))
```

It’s natural to use str_count() with mutate():

```{r mutate - counting number of matchings}
df %>% 
  mutate(
    vowels = str_count(word, "[aeiou]"),
    consonants = str_count(word, "[^aeiou]")
  )
```

Note that matches never overlap. For example, in "abababa", how many times will the pattern "aba" match? Regular expressions say two, not three:

```{r pattern overlap}
str_count("abababa", "aba")
str_view_all("abababa", "aba")
```

##  Extract matches

To extract the actual text of a match, use str_extract(). To show that off, we’re going to need a more complicated example.

```{r sentences}
length(stringr::sentences)
head(stringr::sentences)
```

Imagine we want to find all sentences that contain a colour. We first create a vector of colour names, and then turn it into a single regular expression. Then we can select the sentences that contain a colour, and then extract the colour to figure out which one it is:

```{r matching a vector content}
colours <- c("red", "orange", "yellow", "green", "blue", "purple")
colour_match <- str_c("\\b(", str_c(colours, collapse = "|"), ")\\b")
has_colour <- str_subset(sentences, colour_match)
matches <- str_extract(has_colour, colour_match)
```


Note that str_extract() only extracts the first match. We can see that most easily by first selecting all the sentences that have more than 1 match:

```{r str_extract get only first match}
more <- sentences[str_count(sentences, colour_match) > 1]
str_view_all(more, colour_match)
str_extract(more, colour_match)
```

This is a common pattern for stringr functions, because working with a single match allows you to use much simpler data structures. To get all matches, use str_extract_all(). It returns a list:

```{r str_extract_all to get all matches}
str_extract_all(more, colour_match)
```

If you use simplify = TRUE, str_extract_all() will return a matrix with short matches expanded to the same length as the longest:

```{r str_extract_all results in matrix}
str_extract_all(more, colour_match, simplify = TRUE)
```


## Grouped matches

Earlier in this chapter we talked about the use of parentheses for clarifying precedence and for backreferences when matching. You can also use parentheses to extract parts of a complex match. For example, imagine we want to extract nouns from the sentences. As a heuristic, we’ll look for any word that comes after “a” or “the”. Defining a “word” in a regular expression is a little tricky, so here I use a simple approximation: a sequence of at least one character that isn’t a space.

```{r grouped matches}
noun <- "(a|the) ([^ ]+)"

has_noun <- sentences %>%
  str_subset(noun) %>%
  head(10)

has_noun %>% 
  str_extract(noun)

has_noun %>% 
  str_match(noun)
```

str_extract() gives us the complete match; str_match() gives each individual component. Instead of a character vector, it returns a matrix, with one column for the complete match followed by one column for each group. Like str_extract(), if you want all matches for each string, you’ll need str_match_all().

(Unsurprisingly, our heuristic for detecting nouns is poor, and also picks up adjectives like smooth and parked.)

If your data is in a tibble, it’s often easier to use tidyr::extract(). It works like str_match() but requires you to name the matches, which are then placed in new columns:

```{r tidyr extract}
tibble(sentence = sentences) %>% 
  tidyr::extract(
    sentence, c("article", "noun"), "(a|the) ([^ ]+)", 
    remove = FALSE
  )
```


## Replacing matches

str_replace() and str_replace_all() allow you to replace matches with new strings. The simplest use is to replace a pattern with a fixed string:

```{r replace with string}
x <- c("apple", "pear", "banana")
str_replace(x, "[aeiou]", "-")

str_replace_all(x, "[aeiou]", "-")
```

With str_replace_all() you can perform multiple replacements by supplying a named vector:

```{r replace all with mapping vector}
x <- c("1 house", "2 cars", "3 people")
str_replace_all(x, c("1" = "one", "2" = "two", "3" = "three"))
```

Instead of replacing with a fixed string you can use backreferences to insert components of the match. In the following code, I flip the order of the first three words.

```{r change using backreferences}
head(sentences)

sentences %>% 
  str_replace("([^ ]+) ([^ ]+) ([^ ]+)", "\\3 \\2 \\1") %>% 
  head(5)
```

## Splitting

Use str_split() to split a string up into pieces. For example, we could split sentences into words:

```{r splitting}
sentences %>%
  head(5) %>% 
  str_split(" ")
```

Because each component might contain a different number of pieces, this returns a list. If you’re working with a length-1 vector, the easiest thing is to just extract the first element of the list:

```{r extract first element}
"a|b|c|d" %>% 
  str_split("\\|") %>% 
  .[[1]]
```

Otherwise, like the other stringr functions that return a list, you can use simplify = TRUE to return a matrix:

```{r extract results in a matrix}
sentences %>%
  head(5) %>% 
  str_split(" ", simplify = TRUE)
```

You can also request a maximum number of pieces:

```{r matrix resulting capped number of columns}
fields <- c("Name: Hadley", "Country: NZ", "Age: 35: years old")
fields %>% str_split(": ", n = 2, simplify = TRUE)
```

Instead of splitting up strings by patterns, you can also split up by character, line, sentence and word boundary()s:

```{r boundary() to separate using built-in options}
x <- "This is a sentence.  This is another sentence."
str_view_all(x, boundary("word"))
str_split(x, boundary("sentence"))[[1]]
```

## Find matches

str_locate() and str_locate_all() give you the starting and ending positions of each match. These are particularly useful when none of the other functions does exactly what you want. You can use str_locate() to find the matching pattern, str_sub() to extract and/or modify them.

# Other types of pattern

When you use a pattern that’s a string, it’s automatically wrapped into a call to regex():

```{r regex explicit}
# The regular call:
str_view(fruit, "nana")
# Is shorthand for
str_view(fruit, regex("nana"))
```

You can use the other arguments of regex() to control details of the match:

* ignore_case = TRUE allows characters to match either their uppercase or lowercase forms. This always uses the current locale.

```{r ignore case option}
bananas <- c("banana", "Banana", "BANANA")
str_view(bananas, regex("banana", ignore_case = TRUE))
```

* multiline = TRUE allows ^ and $ to match the start and end of each line rather than the start and end of the complete string.

```{r multiline option}
x <- "Line 1\nLine 2\nLine 3"
str_extract_all(x, "^Line")[[1]]
str_extract_all(x, regex("^Line", multiline = TRUE))[[1]]
```

* comments = TRUE allows you to use comments and white space to make complex regular expressions more understandable. Spaces are ignored, as is everything after #. To match a literal space, you’ll need to escape it: "\\ ".

```{r comments on option}
phone <- regex("
  \\(?     # optional opening parens
  (\\d{3}) # area code
  [) -]?   # optional closing parens, space, or dash
  (\\d{3}) # another three numbers
  [ -]?    # optional space or dash
  (\\d{3}) # three more numbers
  ", comments = TRUE)

str_match("514-791-8141", phone)
```

* dotall = TRUE allows . to match everything, including \n.


# Other uses of regular expressions

There are two useful function in base R that also use regular expressions:

* apropos() searches all objects available from the global environment. This is useful if you can’t quite remember the name of the function.

```{r apropos}
apropos("replace")
```


* dir() lists all the files in a directory. The pattern argument takes a regular expression and only returns file names that match the pattern. For example, you can find all the R Markdown files in the current directory with:

```{r dir regex}
head(dir(pattern = "\\.Rmd$"))
```

(If you’re more comfortable with “globs” like *.Rmd, you can convert them to regular expressions with glob2rx()).

# The stringi package

stringr is built on top of the stringi package. stringr is useful when you’re learning because it exposes a minimal set of functions, which have been carefully picked to handle the most common string manipulation functions. stringi, on the other hand, is designed to be comprehensive. It contains almost every function you might ever need: stringi has 234 functions to stringr’s 46.

If you find yourself struggling to do something in stringr, it’s worth taking a look at stringi. The packages work very similarly, so you should be able to translate your stringr knowledge in a natural way. The main difference is the prefix: str_ vs. stri_.