---
title: "tidy data"
author: "Hernán Estrin"
date: "August 1, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Introduction

In this chapter, you will learn a consistent way to organise your data in R, an organisation called tidy data. Getting your data into this format requires some upfront work, but that work pays off in the long term. Once you have tidy data and the tidy tools provided by packages in the tidyverse, you will spend much less time munging data from one representation to another, allowing you to spend more time on the analytic questions at hand.

This chapter will give you a practical introduction to tidy data and the accompanying tools in the tidyr package.

# Tidy data

You can represent the same underlying data in multiple ways. The example below shows the same data organised in four different ways. Each dataset shows the same values of four variables country, year, population, and cases, but each dataset organises the values in a different way.

```{r tables example}
table1
table2
table3
table4a
table4b
```


These are all representations of the same underlying data, but they are not equally easy to use. One dataset, the tidy dataset, will be much easier to work with inside the tidyverse.

There are three interrelated rules which make a dataset tidy:

* Each variable must have its own column.
* Each observation must have its own row.
* Each value must have its own cell.

These three rules are interrelated because it’s impossible to only satisfy two of the three. That interrelationship leads to an even simpler set of practical instructions:

* Put each dataset in a tibble.
* Put each variable in a column.

In this example, only table1 is tidy. It’s the only representation where each column is a variable.

Why ensure that your data is tidy? There are two main advantages:

* There’s a general advantage to picking one consistent way of storing data. If you have a consistent data structure, it’s easier to learn the tools that work with it because they have an underlying uniformity.

* There’s a specific advantage to placing variables in columns because it allows R’s vectorised nature to shine. As you learned in mutate and summary functions, most built-in R functions work with vectors of values. That makes transforming tidy data feel particularly natural.

*dplyr*, *ggplot2*, and *all the other packages in the tidyverse* are designed to work with tidy data.

# Spreading and gathering

The principles of tidy data seem so obvious that you might wonder if you’ll ever encounter a dataset that isn’t tidy. Unfortunately, however, most data that you will encounter will be untidy. There are two main reasons:

    Most people aren’t familiar with the principles of tidy data, and it’s hard to derive them yourself unless you spend a lot of time working with data.

    Data is often organised to facilitate some use other than analysis. For example, data is often organised to make entry as easy as possible.

This means for most real analyses, you’ll need to do some tidying. The first step is always to figure out what the variables and observations are. Sometimes this is easy; other times you’ll need to consult with the people who originally generated the data. The second step is to resolve one of two common problems:

    One variable might be spread across multiple columns.

    One observation might be scattered across multiple rows.

Typically a dataset will only suffer from one of these problems; it’ll only suffer from both if you’re really unlucky! To fix these problems, you’ll need the two most important functions in tidyr: gather() and spread().

## Gathering

A common problem is a dataset where some of the column names are not names of variables, but values of a variable. Take **table4a**: the column names 1999 and 2000 represent values of the year variable, and each row represents two observations, not one.

To tidy a dataset like this, we need to gather those columns into a new pair of variables. To describe that operation we need three parameters:

* The set of columns that represent values, not variables. In this example, those are the columns 1999 and 2000.

* The name of the variable whose values form the column names. I call that the key, and here it is year.

* The name of the variable whose values are spread over the cells. I call that value, and here it’s the number of cases.

```{r gather example}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```

The columns to gather are specified with dplyr::select() style notation. Here there are only two columns, so we list them individually. Note that “1999” and “2000” are non-syntactic names (because they don’t start with a letter) so we have to surround them in backticks.

In the final result, the gathered columns are dropped, and we get new key and value columns. Otherwise, the relationships between the original variables are preserved. Visually, this is shown in Figure 12.2. We can use gather() to tidy **table4b** in a similar fashion. The only difference is the variable stored in the cell values:

```{r another example of gather} 
table4b %>% 
  gather(`1999`, `2000`, key = "year", value = "population")
```
 
To combine the tidied versions of table4a and table4b into a single tibble, we need to use dplyr::left_join(), which you’ll learn about in relational data.

## Spreading

Spreading is the opposite of gathering. You use it when an observation is scattered across multiple rows. For example, take **table2**: an observation is a country in a year, but each observation is spread across two rows.

To tidy this up, we first analyse the representation in similar way to gather(). This time, however, we only need two parameters:

  * The column that contains variable names, the key column. Here, it’s type.

  * The column that contains values from multiple variables, the value column. Here it’s count.

Once we’ve figured that out, we can use spread(), as shown programmatically below.

```{r spread}
table2 %>%
    spread(key = type, value = count)
```

As you might have guessed from the common key and value arguments, spread() and gather() are complements. gather() makes wide tables narrower and longer; spread() makes long tables shorter and wider.

# Separating and uniting

So far you’ve learned how to tidy table2 and table4, but not table3. table3 has a different problem: we have one column (rate) that contains two variables (cases and population). To fix this problem, we’ll need the *separate()* function. You’ll also learn about the complement of separate(): *unite()*, which you use if a single variable is spread across multiple columns.

## Separate

separate() pulls apart one column into multiple columns, by splitting wherever a separator character appears.

The rate column contains both cases and population variables, and we need to split it into two variables. separate() takes the name of the column to separate, and the names of the columns to separate into.

```{r separate}
table3 %>% 
  separate(rate, into = c("cases", "population"))
```

By default, separate() will split values wherever it sees a non-alphanumeric character (i.e. a character that isn’t a number or letter). For example, in the code above, separate() split the values of rate at the forward slash characters. If you wish to use a specific character to separate a column, you can pass the character to the sep argument of separate().

Look carefully at the column types: you’ll notice that cases and population are character columns. This is the default behaviour in separate(): it leaves the type of the column as is. Here, however, it’s not very useful as those really are numbers. We can ask separate() to try and convert to better types using convert = TRUE.

```{r separate enhanced}
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/", convert = TRUE)
```

You can also pass a vector of integers to sep. separate() will interpret the integers as positions to split at. Positive values start at 1 on the far-left of the strings; negative value start at -1 on the far-right of the strings. When using integers to separate strings, the length of sep should be one less than the number of names in into.

```{r separate by position}
(table5 <- 
  table3 %>% 
  separate(year, into = c("millenium","century","decade", "year"), sep = c(1,2,3,4)))

```

## Unite

unite() is the inverse of separate(): it combines multiple columns into a single column. You’ll need it much less frequently than separate(), but it’s still a useful tool to have in your back pocket.

We can use unite() to rejoin the century and year columns that we created in the last example. That data is saved as tidyr::table5. unite() takes a data frame, the name of the new variable to create, and a set of columns to combine, again specified in dplyr::select() style:

```{r unite}
table5 %>% 
  unite(century, millenium, century, sep = "")
```

# Missing values

Changing the representation of a dataset brings up an important subtlety of missing values. Surprisingly, a value can be missing in one of two possible ways:

* Explicitly, i.e. flagged with NA.
* Implicitly, i.e. simply not present in the data.

Let’s illustrate this idea with a very simple data set:

```{r missing values}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```

There are two missing values in this dataset:

* The return for the fourth quarter of 2015 is explicitly missing, because the cell where its value should be instead contains NA.

* The return for the first quarter of 2016 is implicitly missing, because it simply does not appear in the dataset.

One way to think about the difference is with this Zen-like koan: An explicit missing value is the presence of an absence; an implicit missing value is the absence of a presence.

The way that a dataset is represented can make implicit values explicit. For example, we can make the implicit missing value explicit by putting years in the columns:

```{r expliciting implicit NAs}
stocks %>% 
  spread(year, return)
```

Because these explicit missing values may not be important in other representations of the data, you can set na.rm = TRUE in gather() to turn explicit missing values implicit:

```{r removing missing values}
stocks %>% 
  spread(year, return) %>% 
  gather(year, return, `2015`:`2016`, na.rm = TRUE)
```

Another important tool for making missing values explicit in tidy data is *complete()*:

```{r completing implicit NA}
stocks %>% 
  complete(year, qtr)
```

*complete()* takes a set of columns, and finds all unique combinations. It then ensures the original dataset contains all those values, filling in explicit NAs where necessary.

There’s one other important tool that you should know for working with missing values. Sometimes when a data source has primarily been used for data entry, missing values indicate that the previous value should be carried forward. You can fill in these missing values with *fill()*. It takes a set of columns where you want missing values to be replaced by the most recent non-missing value (sometimes called last observation carried forward).

```{r filling by carrying forward}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)

treatment %>% 
  fill(person)
```


# Non-tidy data

Before we continue on to other topics, it’s worth talking briefly about non-tidy data. Earlier in the chapter, I used the pejorative term “messy” to refer to non-tidy data. That’s an oversimplification: there are lots of useful and well-founded data structures that are not tidy data. There are two main reasons to use other data structures:

* Alternative representations may have substantial performance or space advantages.

* Specialised fields have evolved their own conventions for storing data that may be quite different to the conventions of tidy data.

Either of these reasons means you’ll need something other than a tibble (or data frame). If your data does fit naturally into a rectangular structure composed of observations and variables, I think tidy data should be your default choice. But there are good reasons to use other structures; tidy data is not the only way.

