---
title: "vectors"
author: "Hernán Estrin"
date: "8/17/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Introduction

So far this book has focussed on tibbles and packages that work with them. But as you start to write your own functions, and dig deeper into R, you need to learn about vectors, the objects that underlie tibbles. If you’ve learned R in a more traditional way, you’re probably already familiar with vectors, as most R resources start with vectors and work their way up to tibbles.

Vectors are particularly important as most of the functions you will write will work with vectors. It is possible to write functions that work with tibbles (like ggplot2, dplyr, and tidyr), but the tools you need to write such functions are currently idiosyncratic and immature.

# Vector basics

There are two types of vectors:

* Atomic vectors, of which there are six types: logical, integer, double, character, complex, and raw. Integer and double vectors are collectively known as numeric vectors.

* Lists, which are sometimes called recursive vectors because lists can contain other lists.

The chief difference between atomic vectors and lists is that atomic vectors are *homogeneous*, while lists can be *heterogeneous.* There’s one other related object: NULL. NULL is often used to represent the absence of a vector (as opposed to NA which is used to represent the absence of a value in a vector). NULL typically behaves like a vector of length 0.


Every vector has two key properties:

* Its type, which you can determine with typeof().

```{r}
typeof(letters)
typeof(1:10)
```

* Its length, which you can determine with length().

```{r}
x <- list("a", "b", 1:10)
length(x)
```

Vectors can also contain arbitrary additional metadata in the form of attributes. These attributes are used to create augmented vectors which build on additional behaviour. There are three important types of augmented vector:

* Factors are built on top of integer vectors.
* Dates and date-times are built on top of numeric vectors.
* Data frames and tibbles are built on top of lists.

This chapter will introduce you to these important vectors from simplest to most complicated. You’ll start with atomic vectors, then build up to lists, and finish off with augmented vectors.

# Important types of atomic vector

The four most important types of atomic vector are logical, integer, double, and character. Raw and complex are rarely used during a data analysis, so I won’t discuss them here.

## Logical

Logical vectors are the simplest type of atomic vector because they can take only three possible values: FALSE, TRUE, and NA. Logical vectors are usually constructed with comparison operators, as described in comparisons. You can also create them by hand with c():

```{r logical vectors}
1:10 %% 3 == 0
c(TRUE, TRUE, FALSE, NA)
```

## Numeric

nteger and double vectors are known collectively as numeric vectors. In R, numbers are doubles by default. To make an integer, place an L after the number:

```{r}
typeof(1)
typeof(1L)
```

The distinction between integers and doubles is not usually important, but there are two important differences that you should be aware of:

* Doubles are approximations. Doubles represent floating point numbers that can not always be precisely represented with a fixed amount of memory. This means that you should consider all doubles to be approximations. 

This behaviour is common when working with floating point numbers: most calculations include some approximation error. Instead of comparing floating point numbers using ==, you should use dplyr::near() which allows for some numerical tolerance.

* Integers have one special value: NA, while doubles have four: NA, NaN, Inf and -Inf. All three special values NaN, Inf and -Inf can arise during division:

```{r}
c(-1, 0, 1) / 0
```

Avoid using == to check for these other special values. Instead use the helper functions is.finite(), is.infinite(), and is.nan().

```{r}
is.finite(5)
is.infinite(5L)
is.na(NaN)
is.nan(NA)
```

## Character

Character vectors are the most complex type of atomic vector, because each element of a character vector is a string, and a string can contain an arbitrary amount of data.

You’ve already learned a lot about working with strings in strings.

## Missing values

Note that each type of atomic vector has its own missing value:

```{r}
NA            # logical
NA_integer_   # integer
NA_real_      # double
NA_character_ # character
```
 
Normally you don’t need to know about these different types because you can always use NA and it will be converted to the correct type using the implicit coercion rules described next. However, there are some functions that are strict about their inputs, so it’s useful to have this knowledge sitting in your back pocket so you can be specific when needed.

# Using atomic vectors

Now that you understand the different types of atomic vector, it’s useful to review some of the important tools for working with them. These include:

* How to convert from one type to another, and when that happens automatically.

* How to tell if an object is a specific type of vector.

* What happens when you work with vectors of different lengths.

* How to name the elements of a vector.

* How to pull out elements of interest.

### Coercion

There are two ways to convert, or coerce, one type of vector to another:

* Explicit coercion happens when you call a function like as.logical(), as.integer(), as.double(), or as.character(). Whenever you find yourself using explicit coercion, you should always check whether you can make the fix upstream, so that the vector never had the wrong type in the first place. For example, you may need to tweak your readr col_types specification.

* Implicit coercion happens when you use a vector in a specific context that expects a certain type of vector. For example, when you use a logical vector with a numeric summary function, or when you use a double vector where an integer vector is expected.

Because explicit coercion is used relatively rarely, and is largely easy to understand, I’ll focus on implicit coercion here.

You’ve already seen the most important type of implicit coercion: using a logical vector in a numeric context. In this case TRUE is converted to 1 and FALSE converted to 0. That means the sum of a logical vector is the number of trues, and the mean of a logical vector is the proportion of trues:

```{r logical coercion type}
x <- sample(20, 100, replace = TRUE)
y <- x > 10
sum(y)  # how many are greater than 10?
mean(y) # what proportion are greater than 10?
```

It’s also important to understand what happens when you try and create a vector containing multiple types with c(): the most complex type always wins.

```{r}
typeof(c(TRUE, 1L))
typeof(c(1L, 1.5))
typeof(c(1.5, "a"))
```

An atomic vector can not have a mix of different types because the type is a property of the complete vector, not the individual elements. If you need to mix multiple types in the same vector, you should use a list, which you’ll learn about shortly.


### Test functions

Sometimes you want to do different things based on the type of vector. One option is to use typeof(). Another is to use a test function which returns a TRUE or FALSE. Base R provides many functions like is.vector() and is.atomic(), but they often return surprising results. Instead, it’s safer to use the is_* functions provided by purrr, which are summarised in the table below.

```{r}
is_logical(T)
is_integer(2L)
is_double(2.3)
is_numeric(3L)
is_character("aa")
is_atomic(list(1,2))
is_list(list(1,2))
is_vector(list(1,3))
is_scalar_list(list(1,2))

```

Each predicate also comes with a “scalar” version, like is_scalar_atomic(), which checks that the length is 1. This is useful, for example, if you want to check that an argument to your function is a single logical value.


##  Scalars and recycling rules

As well as implicitly coercing the types of vectors to be compatible, R will also implicitly coerce the length of vectors. This is called vector recycling, because the shorter vector is repeated, or recycled, to the same length as the longer vector.

This is generally most useful when you are mixing vectors and “scalars”. I put scalars in quotes because R doesn’t actually have scalars: instead, a single number is a vector of length 1. Because there are no scalars, most built-in functions are vectorised, meaning that they will operate on a vector of numbers. That’s why, for example, this code works:

```{r recycling vectors}
sample(10) + 100
runif(10) > 0.5
```

In R, basic mathematical operations work with vectors. That means that you should never need to perform explicit iteration when performing simple mathematical computations.

It’s intuitive what should happen if you add two vectors of the same length, or a vector and a “scalar”, but what happens if you add two vectors of different lengths?

```{r}
1:10 + 1:2
```

Here, R will expand the shortest vector to the same length as the longest, so called recycling. This is silent except when the length of the longer is not an integer multiple of the length of the shorter:

```{r}
1:10 + 1:3
```


While vector recycling can be used to create very succinct, clever code, it can also silently conceal problems. For this reason, the vectorised functions in tidyverse will throw errors when you recycle anything other than a scalar. If you do want to recycle, you’ll need to do it yourself with rep():

```{r}
tibble(x = 1:4, y = 1:2)

tibble(x = 1:4, y = rep(1:2, 2))


```

### Naming vectors

All types of vectors can be named. You can name them during creation with c():

```{r naming vectors}
c(x = 1, y = 2, z = 4)
```

Or after the fact with purrr::set_names():

```{r}
set_names(1:3, c("a", "b", "c"))
```

Named vectors are most useful for subsetting, described next.
